---
title: "Earthquake Research"
author: "Leona Pierce, Matt Collins"
code-fold: true
format: pdf
---

## Intro

## Set up

The first step before beginning the investigation of the data is to set up our environment with the necessary packages and import the data. This is done in the code block below.

```{r}
#| label: load-new-packages
#| include: true
#| warning: false
#| echo: true

library(tidyverse)
earthquake = read.csv("./data/earthquake_data.csv")
```

To make the data easier to use, it is also important to tidy the data. In this case, the data itself is tidy, but the column names could be shorter and more direct to make it easier to reference them within the code.

```{r}
#| label: column-rename
#| include: true
#| warning: false
#| echo: true

names(earthquake)[1] <- "general_worry"
names(earthquake)[2] <- "big_one_worry"
names(earthquake)[3] <- "big_one_occur"
names(earthquake)[4] <- "experienced_earthquake"
names(earthquake)[5] <- "taken_preacautions"
names(earthquake)[6] <- "san_adreas_familiar"
names(earthquake)[7] <- "yellowstone_familiar"
names(earthquake)[8] <- "age"
names(earthquake)[9] <- "gender"
names(earthquake)[10] <- "household_income"
names(earthquake)[11] <- "region"

```
## Research Questions
### Question 1
### Question 2
### Question 3
### Question 4
## Conclusion